---
title: "2023_02_23_CTCFfromJun2022"
output: html_document
author: JR with Tenaya's changes
objective: to run John's and practice the pipeline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)
#library(tidyverse)
library(dplyr)
library(readr)
library(Rsubread)
library(rtracklayer)
library(DESeq2)
library(ggpubr)
source("/scratch/Shares/rinn/tenaya/util/_setup.R")
source("/scratch/Shares/rinn/tenaya/util/rinnlab_plot_theme.R")
```


# (1) run macs in bin dir
```{bash}
#!/bin/bash
#SBATCH -p short
#SBATCH --job-name=RPB1_macs
#SBATCH --mail-type=END,FAIL
#SBATCH --mail-user=tenaya.vallery@colorado.edu
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --mem=100gb
#SBATCH --time=6:00:00
#SBATCH --output=RPB1_chip_macs.out
#SBATCH --error=RPB1_chip_macs.err
date; hostname

BASE_PATH=/scratch/Shares/rinn/tenaya/rChIP/rChIP/2022_10_04_RPB1/results/bwa/mergedLibrary

source activate macs3

macs3 callpeak -f BAMPE -t \
${BASE_PATH}/RPB1_NoR_R1.mLb.clN.sorted.bam \
${BASE_PATH}/RPB1_NoR_R2.mLb.clN.sorted.bam \
-c \
${BASE_PATH}/input_R1.mLb.clN.sorted.bam  \
${BASE_PATH}/input_R2.mLb.clN.sorted.bam \
-g hs -n RPB1_chipseq_combined  -q 0.01 --outdir .
date
```


# (2) Create simple annotation file (.saf) for input into FeatureCounts
```{r loading in the consensus peak file (- RNase ) }
# loading in the narrow peak consensus peaks between two CTCF ChIPseq replicates (June 2022)
# Two inputs compared to two replicates and run on narrow peak instead of NF_CORE output broadPeak
rpb1_peaks <- import("/scratch/Shares/rinn/tenaya/rChIP/rChIP/2022_10_04_RPB1/2023_03_14_Attempt2/RPB1_chipseq_combined_peaks.narrowPeak")
# The input to FeatureCounts is .saf (simple annotation file) which will quantify the reads in each peak
# we need a SAF file needs that requires these colnames:
# GeneID		Chr	Start	End	Strand
# creating SAF file below
 # renaming cols to fit format of .saf
# making sure to only grab normal chr not contigs (e.g., GL000009.2)
rpb1_saf <- rpb1_peaks %>%
  as.data.frame() %>%
  filter(grepl("chr", seqnames)) %>%
  dplyr::rename(GeneID = name,
                Chr = seqnames,
                Start = start,
                End = end,
                Strand = strand) %>%
  dplyr::select(c(GeneID, Chr, Start, End, Strand))

# Writting out .saf
write.table(rpb1_saf, "/scratch/Shares/rinn/tenaya/rChIP/rChIP/2022_10_04_RPB1/2023_03_14_Attempt2/rpb1_peaks_combined.saf", sep = "\t", quote = F, row.names = F)
```


# (3) Run feature counts on all bam files +/- rnase
# This will provide quntification of reads in each peak in consensus peaks
# For each sample -- these values can be used as input for deseq below
```{r feature counts across all bam files (+/- RNase)}
# create an object of all bam files (*** including inputs ***)
# these files are far away so creating base-path 
base_path <- "/scratch/Shares/rinn/tenaya/rChIP/rChIP/2022_10_04_RPB1/results/bwa/mergedLibrary/"
# Creating a list of all bams and file paths
rpb1_bams <- c(paste0(base_path, "RPB1_NoR_R1.mLb.clN.sorted.bam"), paste0(base_path, "RPB1_NoR_R2.mLb.clN.sorted.bam"), paste0(base_path, "RPB1_Rnase_R1.mLb.clN.sorted.bam"), paste0(base_path, "RPB1_Rnase_R2.mLb.clN.sorted.bam"), paste0(base_path,"input_R1.mLb.clN.sorted.bam"), paste0(base_path, "input_R2.mLb.clN.sorted.bam"))
# Running feature counts on the bam files from ENCODE that are single ended (-RNase)
# default is single ended but added isPairedEnd parameter to False
rpb1_feature_counts <- featureCounts(files = rpb1_bams,
                                     annot.ext = "/scratch/Shares/rinn/tenaya/rChIP/rChIP/2022_10_04_RPB1/2023_03_14_Attempt2/rpb1_peaks_combined.saf",
                                     isPairedEnd = TRUE,
                                     nthreads = 6)
##Question what does nthreads mean? Says that multiple pathways executed in the program. 1 is default. 

        

# find how many reads in bam file

# saving
write_rds(rpb1_feature_counts, "/scratch/Shares/rinn/tenaya/rChIP/rChIP/2022_10_04_RPB1/2023_03_14_Attempt2/RPB1_feature_counts_output.rds")
```

# start point save and load environment
```{r start point}
# saving env objects:
# save(chip_rnase_data, counts_matrix, ctcf_feature_counts, ctcf_peaks, ctcf_saf, rnase_input_res, rnase_res, chip_input_res, sample_df, samples_chip, samples_rnase, samples_rnase_input, file = "results/ctcf_chip_combined_5_reps/ctcf_rnase.RData")
# load("results/ctcf_chip_combined_5_reps/ctcf_rnase.RData", verbose = T)
# saving minimal objects 
save(rpb1_feature_counts, rpb1_saf, file = "/scratch/Shares/rinn/tenaya/rChIP/rChIP/2022_10_04_RPB1/2023_03_14_Attempt2/minimal_env_object.RData")
# load in 
load("/scratch/Shares/rinn/tenaya/rChIP/rChIP/2022_10_04_RPB1/2023_03_14_Attempt2/minimal_env_object.RData", verbose = T)
```

# (4) creating sample_df for Deseq input
```{r sample sheet for Deseq}
# setting up samples_df data.frame - start with ctcf_feature_counts object
sample_df <-
  data.frame(filename = rpb1_feature_counts[["targets"]]) %>%
  mutate(sample_name = filename)
sample_df$sample_name = gsub(".bam|.mLb.clN.sorted.bam|R1.mLb.clN.sorted.bam",
                             "",
                             sample_df$sample_name) 
# mutate to new col "condition"
# creating sample sheet - file names are in target's col
sample_df <- sample_df %>%
  mutate(condition = ifelse(
    grepl("wash", sample_name),
    "rnase_wash",
    ifelse(
      grepl("NoR", sample_name),
      "rpb1_chip",
      ifelse(grepl("input", sample_name), "input", "rnase")
    )
  ))
```

# (5) create samples and counts for three conditions:
# (i) rnase -vs- chip
```{r creating all the needed count matrix and samples sheets for Deseq}
# getting the counts matrix from feature counts output
counts_matrix <- rpb1_feature_counts$counts
# ChIP +/- comparison
samples_rnase <- sample_df %>%
  filter(condition %in% c("rnase", "rpb1_chip"))
# subsetting counts matrix
rnase_count_matrix <- counts_matrix[, samples_rnase$filename]
# factoring for deseq (+rnase/-rnase)
samples_rnase <- samples_rnase %>%
  mutate(condition = factor(condition, levels = c("rpb1_chip", "rnase")))
# Deseq
dds_rnase <-
  DESeqDataSetFromMatrix(countData = rnase_count_matrix,
                         colData = samples_rnase,
                         design = ~ condition)
# Running Deseq
dds_rnase <- DESeq(dds_rnase)
# getting names
resultsNames(dds_rnase)
# filter out intercept results
rnase_res <-
  results(dds_rnase, name = "condition_rnase_vs_rpb1_chip", tidy = TRUE)
# volcano plot +/- RNase ChIP
ggplot(rnase_res, aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point()
# saving
ggsave("/scratch/Shares/rinn/tenaya/rChIP/rChIP/2022_10_04_RPB1/2023_03_14_Attempt2/chip_plus_minus_rnase_volcano.pdf")
```

# (ii) chip -vs input

```{r chip vs input}
# subsetting chip samples and input
samples_chip <- sample_df %>%
  filter(condition %in% c("input", "rpb1_chip"))
# subsetting counts matrix to chip and input
chip_count_matrix <- counts_matrix[, samples_chip$filename]
# factoring for deseq (chip/input)
samples_chip <- samples_chip %>%
  mutate(condition = factor(condition, levels = c("input", "rpb1_chip")))
# chip Deseq matrix input
dds_chip_input <-
  DESeqDataSetFromMatrix(countData = chip_count_matrix,
                         colData = samples_chip,
                         design = ~ condition)
# running Deseq
dds_chip_input <- DESeq(dds_chip_input)
# getting names
resultsNames(dds_chip_input)
# filter out intercept results
chip_input_res <-
  results(dds_chip_input, name = "condition_rpb1_chip_vs_input", tidy = TRUE)
# volcano plot  ChIP / input
ggplot(chip_input_res, aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point()
# saving
ggsave("/scratch/Shares/rinn/tenaya/rChIP/rChIP/2022_10_04_RPB1/2023_03_14_Attempt2/chip_input_volcano.pdf")
```

# (iii) rnase -vs input
```{r deseq on +rnase chip vs input}
# subsetting to +Rnase chip and input
samples_rnase_input <- sample_df %>%
  filter(condition %in% c("input", "rnase"))
# subsetting counts matrix
rnase_input_count_matrix <-
  counts_matrix[, samples_rnase_input$filename]
# factoring for deseq
samples_rnase_input <- samples_rnase_input %>%
  mutate(condition = factor(condition, levels = c("input", "rnase")))
# setting up deseq +rnase chip/ input
dds_rnase_input <-
  DESeqDataSetFromMatrix(countData = rnase_input_count_matrix,
                         colData = samples_rnase_input,
                         design = ~ condition)
# running Deseq
dds_rnase_input <- DESeq(dds_rnase_input)
# getting names
resultsNames(dds_rnase_input)
# filter out intercept results
rnase_input_res <-
  results(dds_rnase_input, name = "condition_rnase_vs_input", tidy = TRUE)
# volcano plot  +rnase_ChIP / input
ggplot(rnase_input_res, aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point()
```

# ChIP +/- Rnase comparison
# Subselecting Deseq results from three conditions:
(i) RNase-chip -vs- chip
(ii) chip/input
(iii) rnase/input
```{r selecting cols from each result }
# starting with rnase/chip results - renaming and merging together
# with chip/input and rnase/input results.
chip_rnase_data <- rnase_res %>%
  dplyr::select(row, log2FoldChange, pvalue) %>%
  dplyr::rename(l2fc_rnase_vs_chip = log2FoldChange,
                pvalue_rnase_vs_chip = pvalue) %>%
  left_join(chip_input_res %>%
              dplyr::select(row, log2FoldChange, padj) %>%
              dplyr::rename(l2fc_chip_vs_input = log2FoldChange,
                            padj_chip_vs_input = padj)) %>%
  left_join(rnase_input_res %>%
              dplyr::select(row, log2FoldChange, padj) %>%
              dplyr::rename(l2fc_rnase_vs_input = log2FoldChange,
                            padj_rnase_vs_input = padj))
# creating a significant col with mutate
chip_rnase_data <- chip_rnase_data %>%
  mutate(sig = pvalue_rnase_vs_chip <= 0.05)
# chaging row to GeneID to merge with SAF
colnames(chip_rnase_data)[1] <- "GeneID"
# merging with CTCF SAF to get peak coordinates
chip_rnase_data <- left_join(chip_rnase_data, rpb1_saf)
# actually plotting
ggplot(chip_rnase_data, aes(x = l2fc_chip_vs_input, y = l2fc_rnase_vs_input, color = sig)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_abline(slope = 1) +
  geom_smooth(aes(group =1), method = "lm")
# TODO why this doesn't make nice colors :)
theme_paperwhite()
ggsave("/scratch/Shares/rinn/tenaya/rChIP/rChIP/2022_10_04_RPB1/2023_03_14_Attempt2/CTCF_rchip_result.pdf")
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = F)
library(GenomicRanges)
library(dplyr)
library(readr)
# library(Gviz)
library(IRanges)
source("/scratch/Shares/rinn/tenaya/util/intersect_functions.R")
source("/scratch/Shares/rinn/tenaya/util/plotting_functions.R")
library(valr)
```

#promoters
```{r}
# First we need promoters let's make them quick: (actually takes 5 min or so)
gencode_gr <- rtracklayer::import("/scratch/Shares/rinn/genomes/Homo_sapiens/Gencode/v32/gencode.v32.annotation.gtf")
# create promoters 
promoters <- promoters(gencode_gr[gencode_gr$type == "gene"], upstream = 1000, downstream = 1000) 

```



# RBP1
Trying RBP1 as comparison
```{r}
# loading in peak file
RBP1_peaks <- rtracklayer::import("/scratch/Shares/rinn/JR/rchip/data/1-2-2_nextflow_run_all_samples/results/bwa/mergedLibrary/macs/broadPeak/RBP1_no_rnase_R1_peaks.broadPeak")
RBP1_metaplot_df <- data.frame(x = integer(), dens = numeric(), dbp = character())
# running profile TSS 
RBP1_metaplot_df <- profile_tss(RBP1_peaks , promoters)
# plotting metaplot
ggplot(RBP1_metaplot_df, aes(x = x, y = dens)) + 
  geom_line(size = 1.5)
```
#from the Oct 2022 Data - RPB1 No Rnase Replicate 1
```{r}
# loading in peak file
RBP1_peaks <- rtracklayer::import("/scratch/Shares/rinn/tenaya/rChIP/rChIP/2022_10_04_RPB1/results/bwa/mergedLibrary/macs/broadPeak/RPB1_NoR_R1_peaks.broadPeak")
RBP1_metaplot_df <- data.frame(x = integer(), dens = numeric(), dbp = character())
# running profile TSS 
RBP1_metaplot_df <- profile_tss(RBP1_peaks , promoters)
# plotting metaplot
ggplot(RBP1_metaplot_df, aes(x = x, y = dens)) + 
  geom_line(size = 1.5)
```

#from the Oct 2022 Data - RPB1 No Rnase Replicate 2
```{r}
# loading in peak file
RBP1_peaks <- rtracklayer::import("/scratch/Shares/rinn/tenaya/rChIP/rChIP/2022_10_04_RPB1/results/bwa/mergedLibrary/macs/broadPeak/RPB1_NoR_R2_peaks.broadPeak")
RBP1_metaplot_df <- data.frame(x = integer(), dens = numeric(), dbp = character())
# running profile TSS 
RBP1_metaplot_df <- profile_tss(RBP1_peaks , promoters)
# plotting metaplot
ggplot(RBP1_metaplot_df, aes(x = x, y = dens)) + 
  geom_line(size = 1.5)
```