---
title: "20220630_CTCFanalysis"
author: "TV modified from JR"
date: "6/30/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(GenomicRanges)
library(rtracklayer)
library(IRanges)
#library(ComplexHeatmap)
#Could not install ComplexHeatmap because 'rjson' is not available. 
source("/scratch/Shares/rinn/tenaya/rChIP/rChIP/2022_06_15_CTCFconditionstest/intersect_functions.R")
#source("../../util/encode_functions.R")
```

# loading in gencode annotations
```{r gencode annotations}
# loading in gencode annotations
gencode_gr <- rtracklayer::import("/scratch/Shares/rinn/genomes/Homo_sapiens/Gencode/v32/gencode.v32.annotation.gtf")
# creating mRNA and lncRNA promoters (3kb up and downstream)
lncrna_mrna_promoters <- get_promoter_regions(gencode_gr, biotype = c("lncRNA", "protein_coding"))
# export to save for later
rtracklayer::export(lncrna_mrna_promoters, "/scratch/Shares/rinn/JR/rchip/data/gene_annotations/lncRNA_mrna_promoters.gtf")
```



# Finding overlaps between each of the CTCF datasets
# These are two CTCF chip replicates from June 2022 from our lab in K562

```{r determining overlaps between each CTCF chip}
# loading in CTCF new sample 1 peaks
CTCF_10_20 <- rtracklayer::import("/scratch/Shares/rinn/JR/rchip/data/1-2-2_nextflow_run_all_samples/results/bwa/mergedLibrary/macs/broadPeak/CTCF_10-20_shearing_R1_peaks.broadPeak")
# how many peaks?
summary(CTCF_10_20)
# CTCF_1 has 29,365 peaks 
# determining the width of peaks
summary(width(CTCF_10_20))
# Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
# 199.0   297.0   395.0   476.6   543.0 15889.0 
# loading in CTCF new sample 2 peaks
CTCF_20_60 <- rtracklayer::import("/scratch/Shares/rinn/JR/rchip/data/1-2-2_nextflow_run_all_samples/results/bwa/mergedLibrary/macs/broadPeak/CTCF_20-60_shearing_R1_peaks.broadPeak")
# how many peaks
summary(CTCF_20_60)
# CTCF_2 has 34,446 peaks
# determining the width of peaks
summary(width(CTCF_20_60))
# Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
# 170.0   256.0   361.0   449.7   546.0 12775.0
```
### Result: peak number and widths are pretty similar CTCF_2 (20-60) has more peaks [34,446 vs  29, 365]

```{r determining overlaps between each CTCF chip}
# finding overlaps between CTCF_1 & CTCF_2
ov <- findOverlaps(CTCF_10_20, CTCF_20_60)
cv <- countOverlaps(CTCF_10_20, CTCF_20_60)
# number of overlaps?
summary(ov)
# 26,511 overlapping peaks
# determining the overlaps of CTCF_1 in CTCF_2
table(table(ov@to))
##Why choose 'to'?  And not 'from'? Looking ahead it seems like the from and to are the different perspectives from the two datasets. I guess this implies that we can only do 2 datasets at a time. 
table(table(ov@from))
# 26,372 of 29,365 peaks overlap from CTCF_1 to CTCF_2
# 68 peaks had 2 overlpas and 1 peak had 3 overlaps
percent_overlap_1 <- (26372 + (2*68) + 3) / 29365
# 90% overlap
# determining the overlaps of CTCF_2 in CTCF_1
table(table(ov@from))
# 26,215 of 34,446 peaks overlap from CTCF_2 to CTCF_1
# 146 peaks had 2 overlaps and 1 had 3
percent_overlap_2 <- (26215 + (2*146) + 3) / 34446
# 77% of peaks overlap in CTCF_2 in CTCF_1
# trying countoverlaps to compare
#ov_count_overlaps <- countOverlaps(CTCF_10_20, CTCF_20_60)
table(cv)

##Here we see that 3002 do not overlap
```
### Result: there is good overlap betwen the new CTCF samples 90% for CTCF1 and 77% CTCF 2 



# Loading in ENCODE CTCF peaks (replicate 1 and 2 merged from Bernstein group)
```{r loading in ENCODE CTCF IDR replicate 1 and 2 merged narrow peak bed files}
# John Stam CTCF bed narrowPeak	IDR thresholded peaks	1, 2
# https://www.encodeproject.org/experiments/ENCSR000DWE/
stam_encode_ctcf <- read.table("ENCFF736NYC.bed.gz", sep = "\t")
stam_encode_ctcf_gr <- GRanges(seqnames = stam_encode_ctcf$V1,
                     ranges = IRanges(start = stam_encode_ctcf$V2,
                                      end = stam_encode_ctcf$V3))
# Bernstein CTCF bed narrowPeak	IDR thresholded peaks	1, 2
# https://www.encodeproject.org/files/ENCFF362OPG/@@download/ENCFF362OPG.bed.gz
bernstein_encode_ctcf <- read.table("ENCFF362OPG.bed.gz", sep = "\t")
bernstein_encode_ctcf_gr <- GRanges(seqnames = bernstein_encode_ctcf$V1,
                     ranges = IRanges(start = bernstein_encode_ctcf$V2,
                                      end = bernstein_encode_ctcf$V3))
# Snyder CTCF bed narrowPeak	optimal IDR thresholded peaks	1, 2
# https://www.encodeproject.org/files/ENCFF396BZQ/@@download/ENCFF396BZQ.bed.gz
snyder_encode_ctcf <- read.table("ENCFF396BZQ.bed.gz", sep = "\t")
snyder_encode_ctcf_gr <- GRanges(seqnames = snyder_encode_ctcf$V1,
                     ranges = IRanges(start = snyder_encode_ctcf$V2,
                                      end = snyder_encode_ctcf$V3))
# iyer CTCF bed narrowPeak	IDR thresholded peaks	1, 2, 3
# https://www.encodeproject.org/files/ENCFF221SKA/@@download/ENCFF221SKA.bed.gz
iyer_encode_ctcf <- read.table("ENCFF221SKA.bed.gz", sep = "\t")
iyer_encode_ctcf_gr <- GRanges(seqnames = iyer_encode_ctcf$V1,
                     ranges = IRanges(start = iyer_encode_ctcf$V2,
                                      end = iyer_encode_ctcf$V3))
# meyers CTCF bed narrowPeak	optimal IDR thresholded peaks	1, 2
# https://www.encodeproject.org/files/ENCFF119XFJ/@@download/ENCFF119XFJ.bed.gz
meyer_encode_ctcf <- read.table("ENCFF119XFJ.bed.gz", sep = "\t")
meyer_encode_ctcf_gr <- GRanges(seqnames = meyer_encode_ctcf$V1,
                     ranges = IRanges(start = meyer_encode_ctcf$V2,
                                      end = meyer_encode_ctcf$V3))
```


# Finding overlap rates between encode biological replicates to get average overlaps between labs
```{r ENCODE ctcf chip overlaps}
consensus_fl <- list.files("/consensus_peaks", full.names = T)
# using lapply to import all the files in consensus_fl
consensus_peaks <- lapply(consensus_fl, rtracklayer::import)
# Stam versus all
# ctcf_peaks <- GenomicRanges::reduce(CTCF_1, CTCF_2)
table(countOverlaps(stam_encode_ctcf_gr, CTCF_10_20))
##18700 not overlapping and 27460 overlap
length(stam_encode_ctcf_gr)
##total peak count for stam is 46160
stam_ov1 <- findOverlaps(CTCF_10_20, stam_encode_ctcf_gr)
##Reporting 89% overlap between CTCF_10_20 and Stam ENCODE
length(unique(stam_ov1@from))/length(CTCF_10_20)
#93.5% for the reverse
length(unique(stam_ov1@to))/length(CTCF_10_20)
stam_ov2 <- findOverlaps(CTCF_20_60, stam_encode_ctcf_gr)
length(unique(stam_ov2@from))/length(CTCF_20_60)
length(unique(stam_ov2@to))/length(CTCF_20_60)
##The other sample 20/60 is 86% and 89%
# ENCODE 4 PIPELINE -- bed narrowPeak	conservative IDR thresholded peaks	1, 2
bernstein_stam_ov1 <- findOverlaps(stam_encode_ctcf_gr, bernstein_encode_ctcf_gr)
length(unique(bernstein_stam_ov1@from))/length(stam_encode_ctcf_gr)
length(unique(bernstein_stam_ov1@to))/length(stam_encode_ctcf_gr)
##Overlap between stam ENCODE and bernstein ENCODE is 80% and 86%. I'm more than that!!! 

```



# Metaplot of our 2 CTCF chip replicates

```{r Metaplot of June 22 CTCF chip samples}
# This step will accomplish two things: filter out promoters that fall outside the bounds of our coverage vectors
# and filter out chromosomes that are not in common between the promoters object and the peak coverage object. 
# The reason we need to do this is because the peaks may not extend to the end of each chromosome and 
# therefore there is likely to be promoters that fall outside of the peak coverage vectors -- 
# since we know that there are no peaks on those promoters and therefore 
# they don't give us any extra information about where peaks are relative to promoters we'll filter them out.
# Also, it creates problems for the Views command that we'll use to subset the coverage vectors to just the promoter windows.
# First We will use the coverage function to interpret if there is an overlap 
# at given entry of 6,000 entries for each base of the 'promoter window
peak_coverage_CTCF1 <- coverage(CTCF_1)
# This is the length of each run-length encoded vector in the peak_coverage object
# elementNROWs can calculate coverage in RLE
coverage_length <- elementNROWS(peak_coverage_CTCF1)
# we can use the coverage length ranges to 
# filter the promoters falling outside of these boundaries:
# basically turning coverage_lenth to GRanges
coverage_gr <- GRanges(seqnames = names(coverage_length),
                       IRanges(start = rep(1, length(coverage_length)), 
                               end = coverage_length))
# importing promoter annotations in this directory
all_promoters_gr <- rtracklayer::import("/scratch/Shares/rinn/JR/rchip/data/gene_annotations/lncRNA_mrna_promoters.gtf")
# subsetByOverlaps function will find all the promoter ranges within the coverage ranges
# If we didn't do this an error would occur if a chromosome wasn't represented
all_promoters_gr <- subsetByOverlaps(all_promoters_gr, 
                                  coverage_gr, 
                                  type="within", 
                                  ignore.strand=TRUE)
# this is what we are intersecting:
names(peak_coverage)
# with
seqnames(all_promoters_gr)
# note we would have lossed the mitochondrial chromosome!
chromosomes <- intersect(names(peak_coverage_CTCF1), unique(as.character(seqnames(all_promoters_gr))))
# We can also ensure they're in the same order and contain the same chromosomes
# by indexing with this vector
peak_coverage_CTCF1 <- peak_coverage_CTCF1[chromosomes]
# In order to match the list format of the peak_coverage object
# we'll also coerce the GRanges object into an IntegerRangesList.
# If you recall, one of the main features of GRanges object is capturing
# the chromosome information -- when converting to an IRanges list, 
# each chromosome will be represented by a named element in the list.
# using base R here with 'as' a coericing function. 
all_promoters_ir <- as(all_promoters_gr, "IntegerRangesList")[chromosomes]
# ok so Views can store and access all the promoter vectors
# We will end up with an index of all 'promoter windows'
# indexing all the promter window vectors (6Kb)
promoter_peak_view <- Views(peak_coverage_CTCF1, all_promoters_ir)
# This performs functions on each of the vectors in view list.
# kind of like mutate in vector space (note the matrix are ncol and 6,000 rows)
# we will use viewApply to extract if there was a 1 or 0 in the RLE vector for an overlap.
promoter_peak_view <- lapply(promoter_peak_view, function(x) t(viewApply(x, as.vector)))
# We are now going to convert promoter_peak_view to a matrix
# we will do this by rowbinding the Irange list that viewApply
# turned into a vector.
promoter_peak_matrix <- do.call("rbind", promoter_peak_view)
# checking the dimensions
dim(promoter_peak_matrix)
# We're just going to flip one strand because we want to get them in the same orientation
# It really wouldn't matter for a sysmetric window size but good to keep in mind the strand
# could matter.
minus_idx <- which(as.character(strand(all_promoters_gr)) == "-")
# we want to flip so instead of 1:ncol we will do ncol:1 (currenlty 6,000 - 1)
# we want position 1 - 6000
promoter_peak_matrix[minus_idx,] <- promoter_peak_matrix[minus_idx, ncol(promoter_peak_matrix):1]
# Now we can get rid of the rows that have no peaks.
# now we have a matrix of only promoter windows with at least 1 base overlap with
# a promoter region for a given DBP. 
promoter_peak_matrix <- promoter_peak_matrix[rowSums(promoter_peak_matrix) > 0,]
# columns = 6,000 each a base in promoter window.
# rows = all promoter windows.
# sum of cols (colSums) = how many peak overlaps there were at each base (cols).
peak_sums <- colSums(promoter_peak_matrix)
# normalization of binding events to 1 across promoter window
# determining the density of peak coverage at each base in the promoter window
peak_dens <- peak_sums/sum(peak_sums)
# Create a data frame in order to plot this. 
metaplot_df <- data.frame(x = -3e3:(3e3-1), dens = peak_dens)
# Metaplot !
ggplot(metaplot_df, aes(x = x, y = dens)) + 
  geom_line(size = 1.5)
# Nicer Meta Plot
ggplot(metaplot_df, aes(x = x, y = dens)) + 
  geom_vline(xintercept = 0, lty = 2) + 
  geom_line(size = 1.5) + 
  ggtitle("CTCF_1 Promoter Metaplot") + 
  scale_x_continuous(breaks = c(-3000, 0, 3000),
                     labels = c("-3kb", "TSS", "+3kb"),
                     name = "") + 
  ylab("Peak frequency")
```


# Metaplot CTCF 2

```{r metaplot for CTCF 2 sample}
# First We will use the coverage function to interpret if there is an overlap 
# at given entry of 6,000 entries for each base of the 'promoter window
peak_coverage_CTCF2 <- coverage(CTCF_2)
# This is the length of each run-length encoded vector in the peak_coverage object
# elementNROWs can calculate coverage in RLE
coverage_length <- elementNROWS(peak_coverage_CTCF2)
# we can use the coverage length ranges to 
# filter the promoters falling outside of these boundaries:
# basically turning coverage_lenth to GRanges
coverage_gr <- GRanges(seqnames = names(coverage_length),
                       IRanges(start = rep(1, length(coverage_length)), 
                               end = coverage_length))
# creating Granges of coverage vectors that overlap all lncRNA/mRNA promoters
all_promoters_gr <- subsetByOverlaps(all_promoters_gr, 
                                  coverage_gr, 
                                  type="within", 
                                  ignore.strand=TRUE)
# intersecting the seqnames in promoters_gr and coverage names from peak files
chromosomes <- intersect(names(peak_coverage_CTCF2), unique(as.character(seqnames(all_promoters_gr))))
# ordering by chromosomes that overlap
peak_coverage_CTCF2 <- peak_coverage_CTCF2[chromosomes]
# In order to match the list format of the peak_coverage object
# we'll also coerce the GRanges object into an IntegerRangesList.
# If you recall, one of the main features of GRanges object is capturing
# the chromosome information -- when converting to an IRanges list, 
# each chromosome will be represented by a named element in the list.
# using base R here with 'as' a coericing function. 
all_promoters_ir <- as(all_promoters_gr, "IntegerRangesList")[chromosomes]
# This performs functions on each of the vectors in view list.
# kind of like mutate in vector space (note the matrix are ncol and 6,000 rows)
# we will use viewApply to extract if there was a 1 or 0 in the RLE vector for an overlap.
promoter_peak_view <- Views(peak_coverage_CTCF2, all_promoters_ir)
# This performs functions on each of the vectors in view list.
# kind of like mutate in vector space (note the matrix are ncol and 6,000 rows)
# we will use viewApply to extract if there was a 1 or 0 in the RLE vector for an overlap.
promoter_peak_view <- lapply(promoter_peak_view, function(x) t(viewApply(x, as.vector)))
# We are now going to convert promoter_peak_view to a matrix
# we will do this by rowbinding the Irange list that viewApply
# turned into a vector.
promoter_peak_matrix <- do.call("rbind", promoter_peak_view)
# checking the dimensions of the matrix
dim(promoter_peak_matrix)
# We're just going to flip one strand because we want to get them in the same orientation
# It really wouldn't matter for a sysmetric window size but good to keep in mind the strand
# could matter.
minus_idx <- which(as.character(strand(all_promoters_gr)) == "-")
# we want to flip so instead of 1:ncol we will do ncol:1 (currenlty 6,000 - 1)
# we want position 1 - 6000
promoter_peak_matrix[minus_idx,] <- promoter_peak_matrix[minus_idx, ncol(promoter_peak_matrix):1]
# Now we can get rid of the rows that have no peaks.
# now we have a matrix of only promoter windows with at least 1 base overlap with
# a promoter region for a given DBP. 
promoter_peak_matrix <- promoter_peak_matrix[rowSums(promoter_peak_matrix) > 0,]
# columns = 6,000 each a base in promoter window.
# rows = all promoter windows.
# sum of cols (colSums) = how many peak overlaps there were at each base (cols).
peak_sums <- colSums(promoter_peak_matrix)
# determining the density of peak coverage at each base in the promoter window
peak_dens <- peak_sums/sum(peak_sums)
# Create a data frame in order to plot this. 
metaplot_df <- data.frame(x = -3e3:(3e3-1), dens = peak_dens)
# Metaplot
ggplot(metaplot_df, aes(x = x, y = dens)) + 
  geom_line(size = 1.5)
# Nicer metaplot
ggplot(metaplot_df, aes(x = x, y = dens)) + 
  geom_vline(xintercept = 0, lty = 2) + 
  geom_line(size = 1.5) + 
  ggtitle("CTCF_2 Promoter Metaplot") + 
  scale_x_continuous(breaks = c(-3000, 0, 3000),
                     labels = c("-3kb", "TSS", "+3kb"),
                     name = "") + 
  ylab("Peak frequency")
```



# using "encode_file_info" function to get needed information if needed 
```{r encode_file_info function to retrieve data }
# stam_files_df <- encode_file_info("ENCSR000DWE", file_format = "bed") %>%
#   filter(file_format_type == "narrowPeak",
#    output_type == "IDR thresholded peaks",
#    genome_assembly == "GRCh38",
#    biological_replicates == 12) %>%
#   mutate(full_download_url = paste0("https://www.encodeproject.org", download_url))
# system(paste0("cd data/encode_peaks wget ", stam_files_df$full_download_url))
# # https://www.encodeproject.org/files/ENCFF736NYC/@@download/ENCFF736NYC.bed.gz
```
